Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-22 paper.tex
Tidy up OFH section

\subsection{Case study: Genomics England 100,000 genomes}
In this section we demonstrate the utility of VCF Zarr on a large human dataset
and the scalability of the \texttt{vcf2zarr} conversion utility.
Genomics England’s multi-sample VCF dataset (aggV2) is an 
aggregate of 78,195 gVCFs from rare disease and cancer participants 
recruited as part of the 100,000 Genomes Project~\cite{turnbull2018100}. 
The dataset comprises approximately 722 million annotated single-nucleotide 
variants and small indels split into 1,371 roughly equal chunks and 
totalling 165.3 TiB of VCF data after \texttt{bgzip} compression. 
The dataset is used for a variety of research purposes, ranging from 
GWAS~\cite{kousathanas2022whole} and 
imputation~\cite{shi2023genomics} to 
simple queries involving single gene 
regions~\cite{leggatt2023genotype,lam2023repeat}.

@@

\subsection{Case study: Our Future Health genotype release}
Our Future Health~\cite{ofhmain_2025} is the UK’s largest ever health research programme. 
It is designed to help people live healthier lives for longer through the 
discovery and testing of more effective approaches to prevention, earlier 
detection and treatment of diseases.
Our Future Health is inviting millions of people, from all backgrounds and 
from right across the UK to take part. Volunteers are providing information 
about their health and lifestyles to create an incredibly detailed picture 
that represents the whole of the UK.

% this section is probably better placed in acknowledgements?
This study makes use of de-identified data held by Our Future Health. 
We would like to acknowledge all the research participants who have donated 
their data to the Our Future Health research programme 
\url{https://a.storyblok.com/f/228028/x/f45d0da22a/our-future-health-publication-policy-version-1_4.pdf}

Release 9 (\url{https://ourfuturehealth.gitbook.io/our-future-health/data/data-releases/2024-data-releases/release-9}) 
was made available in December 2024. The genetic component includes genotype data 
from 651,050 individuals on 707,522 variants. Here we use the genotype data from 
chromosome 22 to assess the performance of \texttt{vcf2zarr}. The dataset consists of
three VCF files representing chunks of chromosome 22 split by 5,000 variants, totalling 
10,221 variants.

@@

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns produced 
    from the dataset using \texttt{vcf2zarr} with chunk sizes of 1,000 variants and 
    10,000 samples. For column details, see Table~\ref{tab-genomics-england-data}.
    \label{tab-top-fields-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
/call_LAD & int16 & 2.26 GiB & 290.0 & 27.85% \\
/call_GQ & int8 & 1.89 GiB & 87.0 & 23.29% \\
/call_LA & int8 & 1.24 GiB & 260.0 & 15.28% \\
/call_genotype & int8 & 914.22 MiB & 370.0 & 11.00% \\
/call_RGQ & int16 & 729.57 MiB & 460.0 & 8.78% \\
/call_genotype_mask & bool & 606.02 MiB & 550.0 & 7.29% \\
/call_FT & object & 498.61 MiB & 2700.0 & 6.00% \\
\bottomrule
\end{tabular}
\end{table}

@@

Following the conversion process detailed in the previous section, we
first converted the 165 VCF files (7.33TiB) to ICF, followed by
conversion to Zarr. The tasks were run as single jobs on a compute
node with 2,048 GB RAM and 128 cores. Conversion to ICF required
13h45min runtime. The ICF representation used a total of 6.77 TiB over
1,342,569 data storage files. Conversion to Zarr required 18h51min
runtime, generating a dataset with 32 arrays, consuming a total of 6.6
TiB storage over 11,984,581 chunk files. The reduction in total
storage (1.1X) is negligible, mainly due to the inherent difficulty in
compressing the PL field. The top fields in terms of storage are
detailed in Table~\ref{tab-spruce-data}.

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns produced for 
1,063 VCFs on 12 spruce chromosomes using \texttt{vcf2zarr}
default settings. For column details see the caption of Table~\ref{tab-genomics-england-data}.
\label{tab-spruce-data}}
\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {type} & {storage} & {compress} & {\%total} \\
  \midrule
  /call\_PL & int16 & 6.1 TiB & 12.0 & 92.75\% \\
  /call\_genotype & int8 & 282.5 GiB & 26.0 & 4.18\% \\
  /call\_genotype\_mask & bool & 42.2 GiB & 180.0 & 0.62\% \\
  /variant\_DP4 & int32 & 26.2 GiB & 2.1 & 0.39\% \\
  /variant\_MQSBZ & float32 & 13.1 GiB & 1.1 & 0.19\% \\
  /variant\_RPBZ & float32 & 13.0 GiB & 1.1 & 0.19\% \\
  /variant\_MQBZ & float32 & 13.0 GiB & 1.1 & 0.19\% \\
  /variant\_SCBZ & float32 & 12.9 GiB & 1.1 & 0.19\% \\
  /variant\_VDB & float32 & 12.8 GiB & 1.1 & 0.19\% \\
  /variant\_quality & float32 & 12.8 GiB & 1.1 & 0.19\% \\
  /variant\_BQBZ & float32 & 12.6 GiB & 1.1 & 0.19\% \\
  /variant\_SGB & float32 & 12.1 GiB & 1.2 & 0.18\% \\
  /variant\_position & int32 & 9.7 GiB & 1.4 & 0.14\% \\
  /variant\_AC & int16 & 7.4 GiB & 2.8 & 0.11\% \\
  /variant\_DP & int32 & 6.7 GiB & 2.1 & 0.10\% \\
  /variant\_allele & object & 5.4 GiB & 21.0 & 0.08\% \\
  /variant\_AN & int16 & 3.1 GiB & 2.3 & 0.05\% \\
  /call\_genotype\_phased & bool & 2.0 GiB & 1800.0 & 0.03\% \\
  /variant\_filter & bool & 1.5 GiB & 2.4 & 0.02\% \\
  /variant\_MQ & int8 & 489.8 MiB & 7.3 & 0.01\% \\
  \bottomrule
\end{tabular}
\end{table}

Table~\ref{tab-spruce-data} shows that the dataset storage size is
dominated by a major component, call\_PL, which accounts for nearly
93\% of the total. call\_PL is yet another field with low compression
potential due to inherent noisiness.

@@

>>>>

\subsection{Case study: Genomics England WGS data}
In this section we demonstrate the utility of VCF Zarr on a large human dataset
and the scalability of the \texttt{vcf2zarr} conversion utility.
Genomics England’s multi-sample VCF dataset (aggV2) is an 
aggregate of 78,195 gVCFs from rare disease and cancer participants 
recruited as part of the 100,000 Genomes Project~\cite{turnbull2018100}. 
The dataset comprises approximately 722 million annotated single-nucleotide 
variants and small indels split into 1,371 roughly equal chunks and 
totalling 165.3 TiB of VCF data after \texttt{bgzip} compression. 
The dataset is used for a variety of research purposes, ranging from 
GWAS~\cite{kousathanas2022whole} and 
imputation~\cite{shi2023genomics} to 
simple queries involving single gene 
regions~\cite{leggatt2023genotype,lam2023repeat}.

@@

\subsection{Case study: Our Future Health genotype data}
Our Future Health~\cite{ofhmain_2025} is the UK’s largest ever health research programme. 
It is designed to help people live healthier lives for longer through the 
discovery and testing of more effective approaches to prevention, earlier 
detection and treatment of diseases.
Our Future Health is inviting millions of people, from all backgrounds and 
from right across the UK to take part. Volunteers are providing information 
about their health and lifestyles to create an incredibly detailed picture 
that represents the whole of the UK.

Release 9~\citep{ofhdatarelease9}
was made available in December 2024. The genetic component includes genotype data 
from 651,050 individuals on 707,522 variants. Here we use the genotype data from 
chromosome 22 to assess the performance of \texttt{vcf2zarr}. The dataset consists of
three VCF files representing chunks of chromosome 22 split by 5,000 variants, totalling 
10,221 variants.

@@

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns produced 
    from the dataset using \texttt{vcf2zarr} with chunk sizes of 1,000 variants and 
    10,000 samples. For column details, see Table~\ref{tab-genomics-england-data}.
    \label{tab-top-fields-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=4.1]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
% /call_LAD & int16 & 2.26 GiB & 290.0 & 27.85% \\
% /call_GQ & int8 & 1.89 GiB & 87.0 & 23.29% \\
% /call_LA & int8 & 1.24 GiB & 260.0 & 15.28% \\
% /call_genotype & int8 & 914.22 MiB & 370.0 & 11.00% \\
% /call_RGQ & int16 & 729.57 MiB & 460.0 & 8.78% \\
% /call_genotype_mask & bool & 606.02 MiB & 550.0 & 7.29% \\
% /call_FT & object & 498.61 MiB & 2700.0 & 6.00% \\
\bottomrule
\end{tabular}
\end{table}

@@

Following the conversion process detailed in the previous section, we
first converted the 165 VCF files (7.33TiB) to ICF, followed by
conversion to Zarr. The tasks were run as single jobs on a compute
node with 512 GB RAM and 128 cores using 120 cores. Conversion to ICF
required 13h45min runtime. The ICF representation used a total of 6.77
TiB over 1,340,804 data storage files. Conversion to Zarr required
18h36min runtime, generating a dataset with 32 arrays, consuming a
total of 6.6 TiB storage over 10,861,038 chunk files. The reduction in
total storage (1.1X) is negligible, mainly due to the inherent
difficulty in compressing the PL field. The output dataset storage
size is dominated by a major component, call\_PL, which accounts for
nearly 93\% of the total. call\_PL is yet another field with low
compression potential due to inherent noisiness.

Therefore, to reduce storage requirements we generated a Zarr schema
for encoding that uses local allele fields with \texttt{vcf2zarr
mkschema --local-alleles}, and used the generated schema as input to
\texttt{vcf2zarr encode}. Zarr conversion required 14h09min runtime,
generating a dataset with 35 arrays, consuming a total of 2.3TiB
storage over 11,235,556 chunk files, with a reduction in storage of
3.2X. Therefore, by using local allele fields we achieve approximately
3X improvement of compression. The top fields in terms of storage are
detailed inTable~\ref{tab-spruce-data}.

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns
produced for 1,063 VCFs on 12 spruce chromosomes using
\texttt{vcf2zarr} with samples chunk size set to 1,063, variants
chunk size to 10,000, and using local allele fields to reduce output
size. For column details see the caption of
Table~\ref{tab-genomics-england-data}.
\label{tab-spruce-data}}

\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {type} & {storage} & {compress} & {\%total} \\
\midrule
/call\_LPL & int16 & 1.26 T & 17.0 & 54.76\% \\
/call\_LA & int8 & 415.39 G & 18.0 & 17.63\% \\
/call\_genotype & int8 & 381.28 G & 19.0 & 16.18\% \\
/call\_genotype\_mask & bool & 103.84 G & 71.0 & 4.41\% \\
/variant\_DP4 & int32 & 26.24 G & 2.1 & 1.11\% \\
/variant\_MQSBZ & float32 & 13.08 G & 1.1 & 0.56\% \\
/variant\_RPBZ & float32 & 13.04 G & 1.1 & 0.55\% \\
/variant\_MQBZ & float32 & 12.97 G & 1.1 & 0.55\% \\
/variant\_SCBZ & float32 & 12.94 G & 1.1 & 0.55\% \\
/variant\_VDB & float32 & 12.84 G & 1.1 & 0.55\% \\
/variant\_quality & float32 & 12.84 G & 1.1 & 0.55\% \\
/variant\_BQBZ & float32 & 12.64 G & 1.1 & 0.54\% \\
/variant\_SGB & float32 & 12.12 G & 1.2 & 0.51\% \\
/variant\_position & int32 & 9.65 G & 1.4 & 0.41\% \\
/variant\_AC & int16 & 7.42 G & 2.8 & 0.31\% \\
/variant\_DP & int32 & 6.71 G & 2.1 & 0.28\% \\
/variant\_allele & object & 5.37 G & 21.0 & 0.23\% \\
/variant\_AN & int16 & 3.05 G & 2.3 & 0.13\% \\
/call\_genotype\_phased & bool & 1.75 G & 2100.0 & 0.07\% \\
/variant\_filter & bool & 1.46 G & 2.4 & 0.06\% \\
/variant\_MQ & int8 & 489.41 M & 7.3 & 0.02\% \\
/variant\_IMF & float32 & 160.95 M & 89.0 & 0.01\% \\
/variant\_IDV & int16 & 120.72 M & 59.0 & 0.01\% \\
% /variant\_length & int16 & 105.19 M & 68.0 & 0.00\% \\
% /variant\_INDEL & bool & 72.18 M & 49.0 & 0.00\% \\
% /variant\_id & object & 34.53 M & 830.0 & 0.00\% \\
% /variant\_id_mask & bool & 32.74 M & 110.0 & 0.00\% \\
% /variant\_contig & int16 & 32.02 M & 220.0 & 0.00\% \\
% /variant\_FS & float32 & 31.67 M & 450.0 & 0.00\% \\
% /variant\_MQ0F & float32 & 31.67 M & 450.0 & 0.00\% \\
% /region\_index & int32 & 3.14 M & 2.7 & 0.00\% \\
% /contig\_length & int64 & 6.77 K & 1.2 & 0.00\% \\
% /contig\_id & object & 6.2 K & 1.3 & 0.00\% \\
% /sample\_id & object & 5.35 K & 1.6 & 0.00\% \\
% /filter\_id & object & 4.43 K & 0.0 & 0.00\% \\
\bottomrule
\end{tabular}
\end{table}

@@

This study makes use of de-identified data held by Our Future Health. 
We would like to acknowledge all the research participants who have donated 
their data to the Our Future Health research programme~\cite{ofhpubpolicy}.
